A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2019-12-19 paper.Rmd
tidy up up to section 2

For nearly two decades, researchers who work in areas related to computational science have pushed for better standards to verify scientific claims, especially in areas where a full replication of the study would be prohibitively expensive. To meet these minimal standards, there must be easy access to the data, models, and code. Among the different areas with a computational bent, the bioinformatics community in particular has a strong culture around open source [@Gentleman2004], and has made releasing code and associated software a recognized mainstream activity. Many journals in these fields have also pushed authors to submit code (model implementations) alongside their papers, with a few journals going as far as providing a "reproducibility review" [@Peng2011].

@@

Although code sharing in science had a rocky start [@Barnes2010], more and more code writing by scientists is being shared, partly due to the rapid increase in training opportunities made available by organizations like The Carpentries, combined with the rapid adoption of Github by scientists [@Ram2013]. The bigger driver for this may also be connected to the rise in popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has largely been fueled by easy access to open source software tools. Programming languages such as Python, R and Julia help scientists implement and share new methods to work with data. Each of these languages is supported by thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. As of this writing, Python, Julia, and R have  167k packages [@pypi], ~ 14k packages [@cran] and ~2k  packages [@julia-pkgman] respectively. These packages form the building blocks of a data scientists daily workflow. In a modern computational workflow it is trivial for a user to `pip install` a package in Python, or use `install.packages` in R to install all software dependencies for a project. By relying on the idea of having research compendia [@gentleman2007statistical] or something as simple as a requirements file, one can easily install all the necessary scaffolding. Data on the other hard are rarely accessible that easily.

@@

The computing environment is also easily captured with modern tools such as Docker [@Boettiger2015; @Jupyter2018]. Modern tools such as Binder [@Jupyter2018] can parse Docker files and dependency trees to provide on demand, live notebooks in R and Python that a reader can immediately execute in the browser without dealing with the challenges of local installation. This makes it simple to load a specific environment to run any analysis. Code is handled by version control with tools like Git and GitHub [@git; @github], paired with archiving such as Zenodo provide access to code (particularly model implementations)[@zenodo]. All the necessary software is available from various package managers (and their numerous geographic mirrors and archives) making it easy to install any version of a software package. However, the biggest challenge, even today, remains easy and repeatable access to data in a data analysis.

@@

What aspects of data make them particularly challenging to share from a reproducibility perspective? This is the question we tackle in this paper.  While there are several papers that serve as best-practice guides for formatting data [@Broman2017] and getting them ready for sharing [@Ellis2017], the aims of this paper are a bit different. Our aim to address the issue of data in the context of reproducibility in data science workflows. In particular we discuss the various tradeoffs one has to consider when documenting and sharing data, when it is worth publishing and how this would change depending on the use case, perceived impact, and potential audience.

We discuss how to share and/or publish data and cover various tradeoffs when deciding how much to do. We skip detailed discussions of the minutiae of data preparation [@Broman2017; @Ellis2017], or citation (as of this writing, data citation is still in its infancy). We also analyze the state of data contained inside software packages, shared and made available as part of modern data analysis. How often are data shipped inside software packages? How does data format and size impact availability?

# What is the minimal structure to share data for analysis {#minimal}

**Principles for all three**

To share data analysis data, there should be some minimal set of requirements. For example, the data should contain information on metadata, data dictionaries, the README, and data used in analysis (See Section \@ref(minimal)). No matter where data is submitted, there should ideally be a canonical data repository in one long term archive that links to others with a DOI.

@@

Out of these sections, the minimal viable format are:

@@

What goes into each of these sections is now described.

>>>>

For nearly two decades, researchers who work in areas related to computational science have pushed for better standards to verify scientific claims, especially in areas where a full replication of the study would be prohibitively expensive. To meet these minimal standards, there must be easy access to the data, models, and code. Among the different areas with a computational bent, the bioinformatics community in particular has a strong culture around open source [@Gentleman2004], and has made releasing code and associated software a recognized mainstream activity. Many journals in these fields have also pushed authors to submit code (in the form of model implementations) alongside their papers, with a few journals going as far as providing a "reproducibility review" [@Peng2011].

@@

Although code sharing in science had a rocky start [@Barnes2010], more and more code writing by scientists is being shared, partly due to the rapid increase in training opportunities made available by organizations like The Carpentries, combined with the rapid adoption of Github by scientists [@Ram2013]. The bigger driver for this may also be connected to the rise in popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has largely been fueled by easy access to open source software tools. Programming languages such as Python, R and Julia help scientists implement and share new methods to work with data [@rcore; @python; @julia]. Each of these languages is supported by thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. As of this writing, Python, Julia, and R have  167k packages [@pypi], ~ 14k packages [@cran] and ~2k  packages [@julia-pkgman] respectively. These packages form the building blocks of a data scientists daily workflow. In a modern computational workflow it is trivial for a user to `pip install` a package in Python, or use `install.packages` in R to install all software dependencies for a project. By relying on the idea of having research compendia [@gentleman2007statistical] or something as simple as a requirements file, one can easily install all the necessary scaffolding. Data on the other hard are rarely accessible that easily.

@@

The computing environment is also easily captured with modern tools such as Docker [@Boettiger2015; @Jupyter2018]. Modern tools such as Binder [@Jupyter2018] can parse Docker files and dependency trees to provide on demand, live notebooks in R and Python that a reader can immediately execute in the browser without dealing with the challenges of local installation. This makes it simple to load a specific environment to run any analysis. Code is handled by version control with tools like Git and GitHub [@git; @github], paired with archiving frameworks such as Zenodo provide access to code (particularly model implementations)[@zenodo]. All the necessary software is available from various package managers (and their numerous geographic mirrors and archives) making it easy to install any version of a software package. However, the biggest challenge, even today, remains easy and repeatable access to data in a data analysis.

@@

What aspects of data make them particularly challenging to share from a reproducibility perspective? This is the question we tackle in this paper.  While there are several papers that serve as best-practice guides for formatting data [@Broman2017] and getting them ready for sharing [@Ellis2017], the aims of this paper are a bit different. Our aim is to address the issue of data in the context of reproducibility in data science workflows. In particular we discuss the various tradeoffs one has to consider when documenting and sharing data, when it is worth publishing, and how this would change depending on the use case, perceived impact, and potential audience.

We discuss how to share and/or publish data and cover various tradeoffs when deciding how much to do. We skip detailed discussions of data preparation [@Broman2017; @Ellis2017], or citation (as of this writing, data citation is still in its infancy). We also analyze the state of data contained inside software packages, shared and made available as part of modern data analysis. How often are data shipped inside software packages? How does data format and size impact availability?

# A minimal structure to share data for analysis {#minimal}

To share data analysis data, there should be some minimal set of requirements. For example, the data should contain information on metadata, data dictionaries, the README, and data used in analysis (See Section \@ref(minimal)). No matter where data is submitted, there should ideally be a canonical data repository in one long term archive that links to others. It should also have an accession number, or Digital Object Identification (DOI) number, which allows for it to be cited (discussed further in \@ref(cite).

@@

Out of these sections, the minimal viable format are the following provided in one main directory folder:

@@

These sections are now described.
